\thetitle

### 2021-05-22 writing/conclusion.tex
fixing prose

The work of~\citet{tgpk-dls-2018} suggests that, by intuition, programmers
prefer the run-time checking of Natural over other soundness methods. But of course,
the opinion of a random set of programmers does not mean that blame
assignment adds value. Similarly, researchers and language designers have
implicitly answered this question one way or another without evidence for the
blame-strategy dimension. The experiment presented in this paper provides some
justification for the programmers' leanings
\ll{Does it really?}
and helps language creators revisit
their decisions. Of course, the design choice remains a trade-off
along several dimensions while the presented experiment sheds light on
only one of them.

The paper does {\em not\/} address a problem in the gradually typed world that
was pointed out early on by practical researchers~\citep{incorrect-ts,
sta-nt-base-types, wmwz-ecoop-2017} and that has recently received theoretical
attention~\citep{gfd-oopsla-2019, cc-oopsla-20}: mistakes in type annotations
themselves.  Developers use gradual typing to move an untyped code base into the
typed realm, and to this end, they need typed APIs for the vast repositories of
already-existing libraries. Instead of converting the libraries themselves,
language implementors merely create facade modules that import untyped functions
and export them with type annotations, like {\tt typed-pack-lib} in
figure~\ref{fig:example}.  With those facades, the compiler can 
type-check typed modules. But, these retroactive additions of types to a library
may result from a misunderstanding of the code. \emph{Any retroactively ascribed
type may thus be a mistake itself.}

>>>>

The interviews of~\citet{tgpk-dls-2018} suggests that programmers
prefer the run-time checking of Natural over other soundness methods. But,
the opinion of a random set of programmers does not mean that blame
assignment adds value. Similarly, researchers and language designers have
implicitly answered this question one way or another without evidence for the
blame-strategy dimension. The experiment presented in this paper provides some
justification for the programmers' leanings
\ll{Does it really?} \mf{addressed}
and helps language creators revisit
their decisions. Of course, the design choice remains a trade-off
along several dimensions, and  the presented experiment sheds light on
only one of them.

The paper does {\em not\/} address a problem in the gradually typed world that
was pointed out early on by practical researchers~\citep{incorrect-ts,
sta-nt-base-types, wmwz-ecoop-2017} and that has recently received theoretical
attention~\citep{gfd-oopsla-2019, cc-oopsla-20}: mistakes in type annotations
themselves.  Developers use gradual typing to move an untyped code base into the
typed realm, and to this end, they need typed APIs for the vast repositories of
already-existing libraries. Instead of converting the libraries themselves,
language implementors merely create facade modules that import untyped functions
and export them with type annotations, like {\tt typed-pack-lib} in
figure~\ref{fig:example}.  With those facades, the compiler can 
type-check typed modules, but these retroactive additions of types to a library
may result from a misunderstanding of the code. \emph{In short, any retroactively ascribed
type may thus be a mistake itself.}

### 2021-05-22 writing/discussion.tex
fixing prose

The results of the experiment suggests a number of high-level conclusions about
blame strategies in the gradual typed world.  First, run-time type checks have a
large positive impact, regardless of whether these checks assign blame or throw
plain exceptions.  Second, error messages with blame assignments are more
helpful than those without. The results also indicate, though, that blame is not
critical in a majority of cases, and therefore they suggest investigating
whether blame tracking is worth the performance cost.  Third, the Natural
approach fares better than the Transient approach, but only by a small margin.
Since Natural offers complete and sound path-based blame (i.e. it blames all and only those components
that are relevant to a type impedance mismatch) while Transient offers
incomplete but sound heap-based blame~\citep{gfd-oopsla-2019}, the results call
for a study concerning the relative strengths of the two models of blame.
Fourth, given that Transient's sound but shallow run-time type checks do not
seem to hamper debugging, a version of the Natural semantics of Typed Racket
that disables some of its wrappers may offer an answer to the well-known
performance issues of Natural~\citep{gtnffvf-jfp-2019}.  Fifth, the fact that both
modes of the Transient rational programmer are equally successful suggests that
returning the whole blame sequence may not be beneficial. If so, Reticulated
Python could limit the size of blame sequences to attempt to mitigate its
serious performance problems (see below).

@@

Like {\em homo economicus\/}, which decouples the actual behavior of a
participant in an economy for the sake of mathematical modeling, the model of a
rational programmer decouples the actual debugging behavior of a software
developer for the sake of a systematic, large-scale analysis. This decoupling
comes with advantages and disadvantages. In the economic realm, mathematical
models have provided some predictive insights into the market's behavior; but as
behavioral economics has shown more recently (e.g.~\cite{henrich2001search}),
the mathematical abstraction of a
rational actor makes predictions also quite unreliable in some situations.
\footnote{It has misled economists to focus on just mathematics, though
this problem is not relevant here---tongue firmly in cheek.}  Just like an
ordinary consumer or producer, an actual software developer is unlikely to stick
to the exact strategy proposed here. When this happens, the predicted benefits
of blame assignment may not materialize. Indeed, the authors' personal
experience suggests such deviations, and it also suggests that deviating often leads to dead-ends.
To make a true judgment of the usefulness of the rational-programmer
idea, the community will need much more experience with this form of evaluation
and relating the evaluation to the behavior of working programmers.

Relatedly, the experimental setup hides how a rational programmer ascribes types
to extend a trail. When the run-time checks signal an impedance mismatch in the
real world, a real-world programmer does not have a typed module ready to swap
in. Instead, the programmer must come up with the next set of types, which means
making choices. It is usually possible to consistently assign types to variables
in a module in different ways. The maintenance of the benchmarks over many years
has driven home this lesson but, fortunately, it has also shown that the types
are in somewhat canonical.  The authors therefore conjecture that different
real-world programmers would often come up with equivalent type assignments
during debugging sessions.

@@

Section~\ref{sub:mutate-interesting} defines criteria for interesting mutations,
one of which limits the scenarios under consideration to those with mistakes
that raise a run-time error under Erasure. In other words, the experiment is
Erasure-biased: it only considers the usefulness of blame when the safety checks
of the underlying language alone are sufficient to detect the mistake. In
reality, some mistakes require run-time type checks to be
detected~\cite{gfd-oopsla-2019}, and it is possible that blame has more or less
to offer for these kinds of mistakes. If that is the case, then the results of
the experiment on a population of scenarios including such mistakes should be
different.

In fact, the authors have conducted a variation of the experiment that provides
some preliminary evidence that this difference may be significant. In
particular, the variation of the experiment includes only three of the
benchmarks but broadens the selection of scenarios to include any that raise
run-time errors under Natural, regardless of other semantics. The preliminary
results from this experiment suggest that without Erasure-bias, Natural blame
may be much more useful than all of the other modes.

@@

The execution of the experiment reveals that Transient produces empty blame
sequences for 967 scenarios. In theory, empty set must not occur. An empty blame
sequence means a lack of boundary crossings for the witness value.  By
implication, a typed module is blaming itself---something that can happen only
if the type checker (or system) is unsound.

@@

The results in section~\ref{sec:results} also show that the cost of Transient
blame is quite high. Under the Transient semantics, some of the debugging
scenarios exceed the 4-minute timeout or the 6GB-memory limit. To put those
limits into context, the fully typed and fully untyped benchmarks all normally complete
in a few seconds with minimal memory usage. Furthermore, none of the mixed
Natural configurations hit these limits, and with the blame map turned off,
the Transient semantics also runs these programs in a short amount of time and
well within the memory limits. In short, even though the Transient rational
programmer appears to do well in the experiment, the implementation of the
Transient blame strategy might be unrealistic.

At first glance, these measurements seem to contradict the results
of~\citet{vss-popl-2017}. They report an average slowdown of 6.2x and a
worst-case slowdown of 17.2x on the fully-typed Python benchmarks in Reticulated
Python when the blame map gets enabled.  Unfortunately, the average slowdown of 
133x and the
worst-case slowdown of 560x  due to to blame  in Shallow Racket seems closer to the truth.  There are at
least three broad issues that skew the earlier results:
\begin{enumerate}

@@

\item While Reticulated attempts to infer types for local variables, the impoverished nature of its type
    system does not allow the ascription of precise types and often resorts
    to type {\tt Dynamic}~\citep[section~5.4.4]{g-dissertation-2020}. 
    Code with the type {\tt Dynamic}
    has fewer constraints to check at run-time---and much less information
    to track in the blame map.

>>>>

The results of the experiment suggests a number of high-level conclusions about
blame strategies in the gradual typed world.  First, run-time type checks have a
large positive impact, regardless of whether these checks assign blame or throw
plain exceptions.  Second, error messages with blame assignments are more
helpful than those without. The results also indicate, though, that blame is not
critical in a majority of cases, and therefore they suggest investigating
whether blame tracking is worth the performance cost.  Third, the Natural
approach fares better than the Transient approach, but only by a small margin.
Since Natural offers complete and sound path-based blame  while Transient offers
incomplete but sound heap-based blame~\citep{gfd-oopsla-2019}, the results call
for a study concerning the relative strengths of the two models of blame.
Fourth, given that Transient's sound but shallow run-time type checks do not
seem to hamper debugging, a language that supports {\em both\/} Natural and Transient 
might help reduce the number of wrappers and thus address the well-known
performance issues of sound gradual typing~\citep[chapter~6]{g-dissertation-2020}.  Fifth, the fact that both
modes of the Transient rational programmer are equally successful suggests that
returning the whole blame sequence may not be beneficial. If so, Reticulated
Python could limit the size of blame sequences to attempt to mitigate its
serious performance problems (see below).

@@

Like {\em homo economicus\/}, which idealizes the actual behavior of a
participant in an economy for the sake of mathematical modeling, the model of a
rational programmer idealizes the actual debugging behavior of a software
developer for the sake of a systematic, large-scale analysis. This idealization 
comes with advantages and disadvantages. In the economic realm, mathematical
models have provided some predictive insights into the market's behavior; but as
behavioral economics has shown more recently~\cite{henrich2001search},
the mathematical abstraction of a
rational actor makes predictions also quite unreliable in some situations.
\footnote{It has misled economists to focus on just mathematics, though
this problem is not relevant here---tongue firmly in cheek.}  Just like an
ordinary consumer or producer, an actual software developer is unlikely to stick
to the exact strategy proposed here. When this happens, the predicted benefits
of blame assignment may not materialize. Indeed, the authors' personal
experience suggests such deviations, and it also suggests that deviating often leads to dead-ends.
To make a true judgment of the usefulness of the rational-programmer
idea, the community will need much more experience with this form of evaluation
and relating the evaluation to the behavior of working programmers.

Relatedly, the experimental setup hides how a rational programmer ascribes types
to extend a trail. When the run-time checks signal an impedance mismatch in the
real world, a programmer does not have a typed module ready to swap
in. Instead, the programmer must come up with the next set of types, which means
making choices. It is usually possible to consistently assign types to variables
in a module in several different ways. The maintenance of the benchmarks over many years
has driven home this lesson but, fortunately, it has also shown that the types
are in somewhat canonical.  The authors therefore conjecture that different
real-world programmers would often come up with equivalent type assignments
during debugging sessions.

@@

Section~\ref{sub:mutate-interesting} defines criteria for interesting mutations,
one of which limits the scenarios under consideration to those with mistakes
that raise a run-time error under Erasure. In other words, the experiment is
Erasure-biased: it only considers the usefulness of blame when the safety checks
of the underlying language alone are sufficient to detect the mistake. In
reality, some mistakes require run-time type checks to be
detected~\cite{gfd-oopsla-2019}, and it is possible that blame has more 
to offer for these kinds of mistakes. If that is the case, then the results of
the experiment on a population of scenarios including such mistakes should be
different.

In fact, a variation of the experiment provides
some preliminary evidence that this difference may be significant. Thus far,
the variation of the experiment covers only three of the
benchmarks but broadens the selection of scenarios to include any that raise
run-time errors under Natural, regardless of other semantics. The 
results from this small experiment suggest that without Erasure-bias, Natural blame
may be much more useful than all of the other modes.

@@

The execution of the experiment reveals that Transient produces empty blame
sequences for 967 scenarios. An empty blame sequence means a lack of boundary
crossings for the witness value. In theory, an empty sequence should not occur,
because it means a typed module is blaming itself---something that can happen only
if the type checker (or system) is unsound.

@@

The results in section~\ref{sec:results} also show that the cost of Transient
blame is quite high. Under the Transient semantics, some of the debugging
scenarios exceed the 4-minute timeout or the 6GB-memory limit. To put those
limits into context, the fully typed and fully untyped benchmarks all normally complete
in a few seconds with minimal memory usage. Furthermore, none of the mixed
Natural configurations hit these limits, and with the blame map turned off,
the Transient semantics also runs these programs in a short amount of time and
well within the memory limit. In short, even though the Transient rational
programmer appears to do well in the experiment, the implementation of the
Transient blame strategy might be unrealistic.

At first glance, these measurements seem to contradict the results
of~\citet{vss-popl-2017}. They report an average slowdown of 6.2x and a
worst-case slowdown of 17.2x on the fully-typed Python benchmarks in Reticulated
Python when the blame map gets enabled.  Unfortunately, the average slowdown of
133x and the worst-case slowdown of 560x due to blame in Shallow Racket seems
closer to the truth.  There are at least three broad factors that skew Vitousek
et. al.'s results:
\begin{enumerate}

@@

\item While Reticulated attempts to infer types for local variables, the impoverished nature of its type
    system does not allow the ascription of precise types and often resorts
    to type {\tt Dynamic}~\citep[section~5.4.4]{g-dissertation-2020}. 
    Code with type {\tt Dynamic}
    has fewer constraints to check at run-time---and much less information
    to track in the blame map.

### 2021-05-22 writing/rational-natural.tex
fixing prose

\mf{Technically a Natural blame trail could be a Natural exception trail. Hmph? }

Here is a rigorous definition. 
\begin{quote}
\it A \emph{Natural blame trail} is a sequence of scenarios $\conf_0,...\conf_n$ of
a program $\system$ such that for all $0 \leq i \leq n - 1$, $\conf_i \subset
\conf_{i+1}$ and
\[ \conf_{i+1} \setminus \conf_i = %
\left\{ %
\begin{tabular}{ll} %
$\{\blame{\system, \conf_i}\}$      & if $\conf_i$ produces blame \\ %
$\{\exception{\system, \conf_i}\}$  & otherwise \\ %
\end{tabular} \right. %
\]
where $\blame{\system, \conf}$ denotes the module (of $\system$) that $\conf$ blames,
and $\exception{\system, \conf}$ denotes the first untyped location in the stacktrace produced by $\conf$.
\end{quote}

@@

To account for this situation, a new mode of the Natural rational
programmer follows a migration process based entirely on exceptions.
\begin{quote}
\it A {\em Natural exception trail\/} is a sequence of scenarios $\conf_0,...\conf_n$ of a
program $\system$ such that for all $0 \leq i \leq n - 1$, $\conf_i \subset
\conf_{i+1}$ and $\conf_{i+1} \setminus \conf_i = \{\exception{\system, \conf_i}\}$.
\end{quote}

>>>>

\mf{Technically a Natural blame trail could be a Natural exception trail. Hmph? }
\mf{I added a sentence below this definition and the next to address this issue.}

Here is a rigorous definition. 
\begin{quote}
\it A \emph{Natural blame trail} is a sequence of scenarios $\conf_0,...\conf_n$ of
a program $\system$ such that for all $0 \leq i \leq n - 1$, $\conf_i \subset
\conf_{i+1}$ and
\[ \conf_{i+1} \setminus \conf_i = %
\left\{ %
\begin{tabular}{ll} %
$\{\blame{\system, \conf_i}\}$      & if $\conf_i$ produces blame \\ %
$\{\exception{\system, \conf_i}\}$  & otherwise \\ %
\end{tabular} \right. %
\]
where $\blame{\system, \conf}$ denotes the module (of $\system$) that $\conf$ blames,
and $\exception{\system, \conf}$ denotes the first untyped location in the stacktrace produced by $\conf$.
\end{quote}
Note how this definition interprets ``blame information'' broadly, as
any information from a failed run-time check in a blame-oriented semantics.

@@

To account for this situation, a new mode of the Natural rational
programmer follows a migration process based entirely on exceptions.
\begin{quote}
\it A {\em Natural exception trail\/} is a sequence of scenarios $\conf_0,...\conf_n$ of a
program $\system$ such that for all $0 \leq i \leq n - 1$, $\conf_i \subset
\conf_{i+1}$ and $\conf_{i+1} \setminus \conf_i = \{\exception{\system, \conf_i}\}$.
\end{quote}
Using Natural exception trails it becomes possible to factor out
``blame information'' in the narrow sense from the broad one of the
above definition.
